{\plant}: A package for modelling forest trait ecology \& evolution

### 2015-11-09 ms.tex
Some final minor edits

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers
\section{Summary}\label{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
 Population dynamics in forests are strongly size-structured:
 larger plants shade smaller plants while also expending
 proportionately more energy on building and maintaining woody stems.
 Although the importance of size
 structure for demography is widely recognised, many models
 either omit it entirely, or include only coarse approximations.
\item
 Here we introduce the {\plant} package, an
 extensible framework for modelling size- and trait-structured demography,
 ecology and evolution in simulated forests.
 At its core, {\plant} is an
 individual-based model where plant physiology and demography are mediated by
 traits. Individual plants from multiple species can be grown in isolation,
 in patches of competing plants, or in metapopulations under a
 disturbance regime. Patches can be modelled either as a stochastic
 realisation of a finite-sized population, or as their deterministic limit.
 These dynamics can be integrated into metapopulation--level
 estimates of
 invasion fitness and vegetation structure. Because fitness emerges as a
 function of traits, {\plant}
 provides a novel arena for exploring eco-evolutionary dynamics.
\item
 {\plant} is an open source R package and is available at
 \href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
 Accessed from R, the core routines in {\plant} are written in C++.
 The package provides for alternative physiologies and for
 capturing trade-offs among parameters. A
 detailed test suite is provided to ensure correct behaviour of the code.
\item
 {\plant} provides a transparent platform for investigating how
 physiological rules and functional trade-offs interact with competition and
 disturbance regimes to influence vegetation demography, structure and
 diversity.
\end{enumerate}

@@

On top of the physiological model, {\plant} implements methods for
population dynamics and evolutionary dynamics (Fig. \ref{fig:schematic}), following methods
described by \citet{Falster-2011} and \citet{Falster-2015}. Demographic
phenomena can be studied at three levels: individual plants, stands of
competing plants and entire metapopulations. The dynamics at higher
levels of organisation arise as emergent properties, driven by growth
physiology, competition for light and disturbance (Fig. \ref{fig:schematic}).
{\plant} offers the capacity to model emergent phenomena in patches with
specified size, and also through a deterministic approximation
where specifying a patch size is not required. Trait evolution and community
assembly can then be modelled using the estimates of invasion fitness
provided by {\plant}.

@@

While we have implemented a particular physiological model
(\texttt{FF16}), the package is designed to allow arbitrary
additional physiologies to be introduced. Users are able to modify the
physiological model in two different ways. First, one can vary the
parameters of the \texttt{FF16} model directly (see Appendix
\ref{sec:FF16} for the list of parameters and Appendix \ref{sec:code} for code
examples). Second, we allow for changes in key parameters to bring
about changes in other parameters (a hyper-parameterisation), enabling the
straightforward modelling of trade-offs. For example, the
trait \textsc{lma} (leaf mass per unit leaf area) is used by default to estimate
the rate of leaf turnover ($k_{\rm l}$), based on an observed scaling
relationship spanning across diverse vegetation and plant functional
types \citep{Wright-2004},
$$k_{\rm l} = \beta_{kl1} \, \left(\phi / \phi_0\right)^{-\beta_{kl2}},$$
where $k_{\rm l}$ and $\phi$ (\textsc{lma}) are both parameters of the core physiological
model, and $\beta_{kl1}, \beta_{kl2}, \phi_0$ are hyper-parameters. Such
linkages are defined within a user-supplied hyper-parameterisation function
(see Appendix \ref{sec:FF16} for details, or Appendix
\ref{sec:code} for code examples).
%
Finally, rather than varying the parameters of the existing equations
in the \texttt{FF16} model, the equations themselves can be
fundamentally changed. This requires defining a new physiological
model in a new C++ file and compiling that into the package. As an
example, in the \texttt{FFdev} physiology, we have shown how to modify the
function determining allocation to reproduction.

@@

For most applications, however, rather than modelling
dynamics within a finite-sized stochastic patch, it will be preferable to use {\plant}'s
deterministic mode. This assumes that
patches are sufficiently large that population dynamics within a
patch approach their deterministic limit and can
be approximated via a physiologically structured population model \citep{Metz-1986, Deroos-1997, Kohyama-1993}. These models are often formulated as partial differential equations where boundary conditions and coefficients may depend on the population state. Such structured population models are also thought to
capture the average behaviour across a large number of small patches
\citep{Moorcroft-2001}. Importantly, the deterministic mode is
much faster than the stochastic mode, and eliminates the demographic noise
that inevitably arises in finite-sized populations (see Appendix \ref{sec:demography} for details).

Our approach for numerically solving deterministic size-structured population dynamics is
based on the characteristic method \citep{Angulo-2004}. This method describes the development of \(N\) by approximating the
density function along a collection of individual growth trajectories spanning the size
spectrum. Following a disturbance, a series of such trajectories are introduced
into each patch. These trajectories then change according to the growth
of the corresponding individuals, conditioned on their survival
(Fig. \ref{fig:patch}a). Meanwhile, growth and mortality combine
to alter the density of individuals along these trajectories
(Fig. \ref{fig:patch}b). The characteristic method is similar to the Escalator Boxcar Train method
\citep{Deroos-1988, Deroos-1997} used by \citet{Falster-2011} and
\citet{Falster-2015}, but not identical. An advantage of the characteristic method is that it allows direct approximation of the size distribution. For the same reason, however, the characteristic method may be less suitable for resolving size distributions in models where those distributions naturally develop sharp peaks (which happens, e.g., when plants completely cease to grow at a given height).

We implement a novel technique for
handling strongly size-asymmetric competitive feedbacks, such as those occurring
under strong competition for light (see Appendix \ref{sec:demography}
for details). This technique works for both the characteristic method and the Escalator Boxcar Train method. It involves an
adaptive refinement of the time points where new growth trajectories or cohorts are introduced
into the population, an idea first applied in \citet{Falster-2011} and
described further in \citet{Falster-2015}.
Normally, growth trajectories are introduced evenly, spaced out according to a fixed time
interval. Under size-asymmetric competition, however, the growth
trajectories of individuals born at nearby times can diverge
substantially over time (Fig. \ref{fig:patch}a; Appendix
\ref{sec:demography} and \ref{sec:code}). This can become a problem during patch
development, because much of the population can be located in between
widely-spaced trajectories, causing low numerical accuracy. To maintain
numerical accuracy, we use an iterative algorithm to adaptively refine
the cohort spacing until adjacent trajectories remain
adequately resolved throughout patch development (see Appendix
\ref{sec:demography} for details). This refinement gives rise to an uneven
distribution of cohort introduction times across patch age (Fig.
\ref{fig:patch}a), with a considerably tighter spacing of trajectories in
younger patches.

Modelling the full size-structured dynamics within patches enables multiple demographic phenomena to be
investigated. First, {\plant} can simulate demographic
phenomena typically observed in patches
recovering from a disturbance, including self-thinning and
successional replacement. Whereas other approaches for modelling
self-thinning are limited to a single species
\citep[e.g.,][]{Barnes-2004, Coomes-2007}, \textsc{TSPMs} can accommodate
multiple species with different traits (Fig. \ref{fig:schematic}b;
Fig. \ref{fig:patch}). Second, the self-thinning and
successional replacement emerges from the
combined effects of physiology and competition for resources, rather
than being prescribed by the model. As a consequence, it is possible to ask how alternative trait values affect fitness (see below). Third, \textsc{TSPMs} allow the combined effects of physiology and succession on a patch's emergent
properties to be investigated
\citep{Moorcroft-2001, Falster-2011}. In particular, leaf area cover and rates of biomass
production tend to vary nonlinearly with patch
age \citep{Smith-2001, Binkley-2002, Ogawa-2010};
in {\plant}, such outcomes naturally arise as emergent properties,
driven by demography (Fig. \ref{fig:patch}c).

@@

The main numerical challenge in stepping from a single patch to a
metapopulation is to identify the equilibrium seed rain of the
metapopulation. An input seed rain is required to simulate the
demography of the metapopulation, and this in turn produces an output
seed rain. Equilibrium occurs when, for each species,
the input seed rain equals the output seed rain. Identifying this equilibrium
requires finding the root of a multi-dimensional function, subject to
the constraint that a unique positive root exists and the root is dynamically stable (the trivial equilibrium
of zero seed rain always exists, but is often not stable; Appendix
\ref{sec:code}). Solving for multi-dimensional roots can be computationally challenging,
with no single technique guaranteed to find a solution. In {\plant}, we sequentially
applying rounds of iteration and nonlinear root finding algorithms, while also
using some heuristics to try to ensure that the root returned is an attracting point
(see Appendix \ref{sec:demography} for details). Real seed rains are of course noisy,
but the aforementioned approach allows for mean tendencies to be estimated (which is required for computing fitness; see below).

@@

Finally, with the extension to the metapopulation level, we are
modelling demography across the entire plant life cycle and are thus in a
position to estimate invasion fitness -- the rate at which a rare
individual with particular traits can establish (positive invasion fitness) or cannot establish (negative invasion fitness) in a community of
established residents at their equilibrium densities
\citep{Metz-1992}. While often calculated as the
long-term per capita growth rate, a more convenient indicator of
fitness in structured metapopulations is given by the
logarithm of the basic reproduction ratio, \(R\) \citep{Gyllenberg-2001, Metz-2001}. The basic reproduction ratio is simply the average number of seeds produced from one seed of a new type over its lifetime, averaged across the metapopulation. So a new type can
invade when \(R > 1\). The calculation of \(R\) available in
{\plant} follows equations given by \citet{Falster-2015} (
see Appendix \ref{sec:demography} for details).

With this measure of fitness, we can model trait-based adaptive
dynamics, and community assembly \citep{Metz-1992, Geritz-1998,
Chesson-2000, Dieckmann-2004, Brannstrom-2013b}. An example
for forests was introduced by
\citet{Falster-2015}. This is best thought of as occurring on a fitness landscape,
a surface with fitness plotted against one or more traits
(Fig. \ref{fig:fitness}). The reproductive success of individuals
depends both on their own traits and on the densities of other occupants in the current resident community. Fitness landscapes are positive in trait regions that allow invasion, and zero
for resident species that are at equilibrium. The slope of the fitness
landscape with respect to traits determines the direction of selection
(dashed line in Fig. \ref{fig:fitness}a), i.e., a negative slope indicates selection for
smaller trait values (see Appendix \ref{sec:code} for
details). Fitness landscapes are continually reshaped as species are
introduced and their trait values evolve, reflecting the nonlinear demographic feedbacks
that occur under frequency-dependent evolution. Stable coexistence is
possible when species with different traits can mutually invade.
Following the fitness gradient may lead to an
evolutionary branching point; at
such a point, there is no directional selection (solid line in
Fig. \ref{fig:fitness}a), but both smaller and larger trait values can invade.
It is often possible to find a species configuration where
fitness is negative everywhere (solid line in Fig. \ref{fig:fitness}b)
except for the zero points where species are present. Such communities
are immune to invasions by species with any other trait values and represent
predictions for the trait mixtures favoured by
natural selection .

@@

Increased interest in capturing the dynamics of size-structured populations
has prompted development of several tools, each with its own distinct
strengths. The \textsc{EBT} software (see De Roos 2016 in this special feature) provides
a flexible framework that has been used to study a range of marine organisms.
Compared to the \textsc{EBT}, {\plant} is customised for plants, and
to that end offers refinements needed to handle modelling of the strongly
size-asymmetric feedbacks that occur in plant communities, as well as the
extension to a structured metapopulation. Other \textsc{TSPMs} also exist for modelling of vegetation, like \textsc{ED}
\citep[ver 1 and 2;][]{Moorcroft-2001, Medvigy-2009}, \textsc {LPJ-GUESS}
\citep{Smith-2014} and \textsc{LPJmL-FIT} \citep{Sakschewski-2015}. These
models were designed for simulating bio-geochemical cycles and vegetation
structure across broad geographic areas, and their architectures are
optimised for such use. By contrast, {\plant} was is designed so that the
physiological rules driving the dynamics of the system could be easily
manipulated and emergent demographic properties (including those of individual plants, patches, metapopulations of patches and fitness) studied in
detail. Finally, like {\plant}, the \textsc{LM3-PPA} \cite{Weng-2015} model also aims
to scale from individual-level processes to emergent properties of
vegetation, including estimating invasion fitness. Yet the approaches taken
for solving the dynamics are very different. The \textsc{LM3-PPA} relies on
complex analytical solutions to the size-structured interactions, whose
derivation required simplifying elements of the physiology and ecology in that
model. By contrast, in {\plant} we have retained the full range of
physiological and ecological feedbacks, and instead relied on using robust
numerical methods to solve the system dynamics (see Appendix
\ref{sec:demography} for details).

@@

The {\plant} package is designed to make it easier for researchers and forest
practitioners alike to investigate a variety of nonlinear demographic, ecological and evolutionary
phenomena. Rather than provide a model with a few entry points, we
have developed an extensible framework that we hope can be used to
answer a number of questions at a variety of scales. The specific
applications highlighted here arise from our interest in the assembly of
trait mixtures in communities. A variety of other uses are
possible, addressing other aspects of community assembly. For
example, we expect the package will prove useful for i) investigating
how morphological and physiological traits influence growth rate and
shade tolerance, ii) studying emergent phenomena such as multi-species self-thinning, successional transitions and age-related changes in
vegetation productivity, and iii) bridging between empirical data,
large-scale simulators of global biogeochemical cycles and simple
abstract theoretical models.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/patch.pdf}
\caption{Population dynamics for two species competing within a patch.
 Red and blue lines refer to the low- and high-\textsc{lma} species from
 Fig. \ref{fig:plant}. (a) Growth trajectories
 for individuals germinating in successive ``cohorts'' for each
 species, using a schedule of introduction times generated by
 {\plant}'s adaptive algorithm. The initial growth rate advantage of
 the low-\textsc{lma} species (Figure \ref{fig:plant}c) means it
 quickly overtops the high-\textsc{lma} species, suppressing its
 growth. Self-thinning then creates space for the high-\textsc{lma}
 species to establish below the canopy of the faster-growing species.
 (b) Same trajectories as in (a), but shaded
 according to the density $N$ of individuals at given size and patch
 age (light = low density, dark = high density). (c) Leaf area
 index at ground level for each species. The dotted line is the
 total leaf area index (summed across both species), while the lines on
 the right-hand side are averages integrated over the whole
 metapopulation. See Appendix \ref{sec:code} for code
 reproducing this figure.}
\label{fig:patch}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/fitness.pdf}
\caption{Example of a fitness landscape for the \textsc{lma} (leaf mass per
 unit leaf area) trait, showing potential for the stable coexistence of
 multiple plant types \citep[adapted from][]{Falster-2015}. Invasion fitness is the logarithm of the population growth rate --
 at zero fitness, species are at equilibrium, at positive fitness, they increase in
 density. (a) Fitness landscape generated by a single species (indicated by
 a circle). For an \textsc{lma} of 0.08 (solid line, filled circle), there is
 directional selection for larger \textsc{lma}. For an \textsc{lma} of
 0.1 (dashed line, open circle), there is a branching point (see inset). (b) Fitness landscape generated by two species, holding
 the first at an \textsc{lma} of 0.08. Introducing a species at the
 maximum of the solid line in (a) reduces the region of positive
 fitness considerably (dashed line, open circle). Dotted lines show
 subsequent invasions and replacements of the second species. The
 solid line shows the fitness landscape of the resultant evolutionarily stable two-species community, with both species coexisting
 on separate peaks. At this point, no further
 invasion is possible. See Appendix \ref{sec:code} for code
 reproducing this figure.}
\label{fig:fitness}
\end{figure}

>>>>

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers
\section{Summary}\label{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
 Population dynamics in forests are strongly size-structured:
 larger plants shade smaller plants while also expending
 proportionately more energy on building and maintaining woody stems.
 Although the importance of size
 structure for demography is widely recognised, many models
 either omit it entirely, or include only coarse approximations.
\item
 Here we introduce the {\plant} package, an
 extensible framework for modelling size- and trait-structured demography,
 ecology and evolution in simulated forests.
 At its core, {\plant} is an
 individual-based model where plant physiology and demography are mediated by
 traits. Individual plants from multiple species can be grown in isolation,
 in patches of competing plants, or in metapopulations under a
 disturbance regime.
 These dynamics can be integrated into metapopulation--level
 estimates of
 invasion fitness and vegetation structure. Because fitness emerges as a
 function of traits, {\plant}
 provides a novel arena for exploring eco-evolutionary dynamics.
\item
 {\plant} is an open source R package and is available at
 \href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
 Accessed from R, the core routines in {\plant} are written in C++.
 The package provides for alternative physiologies and for
 capturing trade-offs among parameters. A
 detailed test suite is provided to ensure correct behaviour of the code.
\item
 {\plant} provides a transparent platform for investigating how
 physiological rules and functional trade-offs interact with competition and
 disturbance regimes to influence vegetation demography, structure and
 diversity.
\end{enumerate}

@@

On top of the physiological model, {\plant} implements methods for
population dynamics and adaptive dynamics (Fig. \ref{fig:schematic}), following methods
described by \citet{Falster-2011} and \citet{Falster-2015}. Demographic
phenomena can be studied at three levels: individual plants, stands of
competing plants and entire metapopulations. The dynamics at higher
levels of organisation arise as emergent properties, driven by growth
physiology, competition for light and disturbance (Fig. \ref{fig:schematic}).
{\plant} offers the capacity to model emergent phenomena in patches with
specified area, and also through a deterministic approximation
where specifying a patch size is not required. Trait evolution and community
assembly can then be modelled using the estimates of invasion fitness
provided by {\plant}.

@@

While we have implemented a particular physiological model
(\texttt{FF16}), the package is designed to allow arbitrary
additional physiologies to be introduced. Users are able to modify the
physiological model in three different ways. First, one can vary the
parameters of the \texttt{FF16} model directly (see Appendix
\ref{sec:FF16} for the list of parameters and Appendix \ref{sec:code} for code
examples). Second, we allow for changes in key parameters to bring
about changes in other parameters (a hyper-parameterisation), enabling the
straightforward modelling of trade-offs. For example, the
trait \textsc{lma} (leaf mass per unit leaf area) is used by default to estimate
the rate of leaf turnover ($k_{\rm l}$), based on an observed scaling
relationship spanning across diverse vegetation and plant functional
types \citep{Wright-2004},
$$k_{\rm l} = \beta_{{\rm kl}1} \, \left(\phi / \phi_0\right)^{-\beta_{{\rm kl}2}},$$
where $k_{\rm l}$ and $\phi$ (\textsc{lma}) are both parameters of the core physiological
model, and $\beta_{{\rm kl}1}, \beta_{{\rm kl}2}, \phi_0$ are hyper-parameters. Such
linkages are defined within a user-supplied hyper-parameterisation function
(see Appendix \ref{sec:FF16} for details, or Appendix
\ref{sec:code} for code examples).
%
Finally, rather than varying the parameters of the existing equations
in the \texttt{FF16} model, the equations themselves can be
fundamentally changed. This requires defining a new physiological
model in a new C++ file and compiling that into the package. As an
example, in the \texttt{FFdev} physiology, we have shown how to modify the
function determining allocation to reproduction.

@@

For most applications, however, rather than modelling
dynamics within a finite-sized stochastic patch, it will be preferable to use {\plant}'s
deterministic mode. This assumes that
patches are sufficiently large that population dynamics within a
patch approach their deterministic limit and can
be approximated via a physiologically structured population model \citep{Metz-1986, Deroos-1997, Kohyama-1993}. These models are often formulated as partial differential equations where boundary conditions and coefficients may depend on the population state. Such structured population models are also thought to
capture the average behaviour across a large number of small patches
\citep{Moorcroft-2001}. Importantly, the deterministic mode in {\plant} is
much faster than the stochastic mode, and eliminates the demographic noise
that inevitably arises in finite-sized populations (see Appendix \ref{sec:demography} for details).

Our approach for numerically solving deterministic size-structured population dynamics is
based on the characteristic method \citep{Angulo-2004}. This method describes the development of
the size distribution\(N\) by approximating it along a collection of individual growth trajectories spanning the size
spectrum. Following a disturbance, a series of such trajectories are introduced
into each patch. These trajectories then change according to the growth
of the corresponding individuals, conditioned on their survival
(Fig. \ref{fig:patch}a). Meanwhile, growth and mortality combine
to alter the density of individuals along these trajectories
(Fig. \ref{fig:patch}b). The characteristic method is similar to the Escalator Boxcar Train (\textsc{EBT})method
\citep{Deroos-1988, Deroos-1997} used by \citet{Falster-2011} and
\citet{Falster-2015}, but not identical. An advantage of the characteristic method is that it allows direct approximation of the size distribution. For the same reason, however, the characteristic method may be less suitable for resolving size distributions in models where those distributions naturally develop sharp peaks (which happens, e.g., when plants completely cease to grow at a given height).

We implement a novel technique for
handling strongly size-asymmetric competitive feedbacks, such as those occurring
under strong competition for light (see Appendix \ref{sec:demography}
for details). This technique works for both the characteristic method and the \textsc{EBT} method. It involves an
adaptive refinement of the time points where new growth trajectories or cohorts are introduced
into the population, an idea first applied in \citet{Falster-2011} and
described further in \citet{Falster-2015}.
Normally, growth trajectories are introduced evenly, spaced out according to a fixed time
interval. Under size-asymmetric competition, however, the growth
trajectories of individuals born at nearby times can diverge
substantially over time (Fig. \ref{fig:patch}a; Appendix
\ref{sec:demography} and \ref{sec:code}). This can become a problem during patch
development, because much of the population can be located in between
widely-spaced trajectories, causing low numerical accuracy. To maintain
numerical accuracy, we use an iterative algorithm to adaptively refine
the cohort spacing until adjacent trajectories remain
adequately resolved throughout patch development (see Appendix
\ref{sec:demography} for details). This refinement gives rise to an uneven
distribution of cohort introduction times across patch age (Fig.
\ref{fig:patch}a), with a considerably tighter spacing of trajectories in
younger patches.

Modelling the full size-structured dynamics within patches enables multiple demographic phenomena to be
investigated. First, {\plant} can simulate demographic
phenomena typically observed in patches
recovering from a disturbance, including self-thinning and
successional replacement. Whereas other approaches for modelling
self-thinning are limited to a single species
\citep[e.g.,][]{Barnes-2004, Coomes-2007}, \textsc{TSPMs} can accommodate
multiple species with different traits (Fig. \ref{fig:schematic}b;
Fig. \ref{fig:patch}). Second, the self-thinning and
successional replacement emerges from the
combined effects of physiology and competition for resources, rather
than being prescribed by the model. As a consequence, it is possible to ask how alternative trait values affect fitness (see below). Third, \textsc{TSPMs} allow the combined effects of physiology and succession on a patch's emergent
properties to be investigated
\citep{Moorcroft-2001, Falster-2011}. In particular, leaf area cover and rates of biomass
production tend to vary nonlinearly with patch
age \citep{Smith-2001, Binkley-2002, Ogawa-2010};
in {\plant}, such outcomes naturally arise as emergent properties,
driven by underlying physiology and demography (Fig. \ref{fig:patch}c).

@@

The main numerical challenge in stepping from a single patch to a
metapopulation is to identify the equilibrium seed rain of the
metapopulation. An input seed rain is required to simulate the
demography of the metapopulation, and this in turn produces an output
seed rain. Equilibrium occurs when, for each species,
the input seed rain equals the output seed rain. Identifying this equilibrium
requires finding the root of a multi-dimensional function, subject to
the constraint that a unique positive root exists and the root is dynamically stable (the trivial equilibrium
of zero seed rain always exists, but is often not stable; Appendix
\ref{sec:code}). Solving for multi-dimensional roots can be computationally challenging,
with no single technique guaranteed to find a solution. In {\plant}, we sequentially
apply rounds of iteration and nonlinear root finding, while also
using some heuristics to try to ensure that the root returned is an attracting point
(see Appendix \ref{sec:demography}, \ref{sec:code} for details). Real seed rains are of course noisy,
but the aforementioned approach allows for mean tendencies to be estimated (which is required for computing fitness; see below).

@@

Finally, with the extension to the metapopulation level, we are
modelling demography across the entire plant life cycle and are thus in a
position to estimate invasion fitness -- the rate at which a rare
individual with particular traits can establish (positive invasion fitness) or cannot establish (negative invasion fitness) in a community of
established residents at their equilibrium densities
\citep{Metz-1992}. While often calculated as the
long-term per capita growth rate, a more convenient indicator of
fitness in structured metapopulations is given by the
logarithm of the basic reproduction ratio, \(R\) \citep{Gyllenberg-2001, Metz-2001}. The basic reproduction ratio is simply the average number of seeds produced from one seed of a new type over its lifetime, averaged across the metapopulation. So a new type can
invade when \(R > 1\). The calculation of \(R\) available in
{\plant} follows equations given by \citet{Falster-2015}
(see Appendix \ref{sec:demography} for details).

With this measure of fitness, we can model trait-based adaptive
dynamics, and community assembly \citep{Metz-1992, Geritz-1998,
Chesson-2000, Dieckmann-2004, Brannstrom-2013b}. An example
for forests was introduced by
\citet{Falster-2015}. This is best thought of as occurring on a fitness landscape,
a surface with fitness plotted against one or more traits
(Fig. \ref{fig:fitness}). The reproductive success of individuals
depends both on their own traits and on the densities of other occupants in the current resident community. Fitness landscapes are positive in trait regions that allow invasion, and zero
for resident species that are at equilibrium. The slope of the fitness
landscape with respect to traits determines the direction of selection,
for example the negative slope of the dashed line around the resident
in Fig. \ref{fig:fitness}a)indicates selection for
smaller trait values (see Appendix \ref{sec:code} for
details). Fitness landscapes are continually reshaped as species are
introduced and their trait values evolve, reflecting the nonlinear demographic feedbacks
that occur under frequency-dependent evolution. Stable coexistence is
possible when species with different traits can mutually invade.
Following the fitness gradient may lead to an
evolutionary branching point; at
such a point, there is no directional selection (solid line in
Fig. \ref{fig:fitness}a), but both smaller and larger trait values can invade.
It is often possible to find a species configuration where
fitness is negative everywhere (solid line in Fig. \ref{fig:fitness}b)
except for the zero points where species are present. Such communities
are immune to invasions by species with any other trait values and represent
predictions for the trait mixtures favoured by
natural selection .

@@

Increased interest in capturing the dynamics of size-structured populations
has prompted development of several tools, each with its own distinct
strengths. The \textsc{EBT} software (see De Roos 2016 in this special feature) provides
a flexible framework that has been used to study a range of organisms.
Compared to the \textsc{EBT}, {\plant} is customised for plants, and
to that end offers refinements needed to handle modelling of the strongly
size-asymmetric feedbacks that occur in plant communities, as well as the
extension to a structured metapopulation. Other \textsc{TSPMs} also exist for modelling of vegetation, like \textsc{ED}
\citep[ver 1 and 2;][]{Moorcroft-2001, Medvigy-2009}, \textsc {LPJ-GUESS}
\citep{Smith-2014} and \textsc{LPJmL-FIT} \citep{Sakschewski-2015}. These
models were designed for simulating bio-geochemical cycles and vegetation
structure across broad geographic areas, and their architectures are
optimised for such use. By contrast, {\plant} was designed so that the
physiological rules driving the dynamics of the system could be easily
manipulated and emergent demographic properties (including those of individual plants, patches, metapopulations of patches and fitness) studied in
detail. Finally, like {\plant}, the \textsc{LM3-PPA} \cite{Weng-2015} model also aims
to scale from individual-level processes to emergent properties of
vegetation, including estimating invasion fitness. Yet the approaches taken
for solving the dynamics are very different. The \textsc{LM3-PPA} relies on
complex analytical solutions to the size-structured interactions, whose
derivation required simplifying elements of the physiology and ecology in that
model. By contrast, in {\plant} we have retained the full range of
physiological and ecological feedbacks, and instead relied on using robust
numerical methods to solve the system dynamics (see Appendix
\ref{sec:demography} for details).

@@

The {\plant} package is designed to make it easier for researchers and forest
practitioners alike to investigate a variety of nonlinear demographic, ecological and evolutionary
phenomena. Rather than provide a model with a few entry points, we
have developed an extensible framework that we hope can be used to
answer a number of questions at a variety of scales. The specific
applications highlighted here arise from our interest in the assembly of
trait mixtures in communities. A variety of other uses are
possible, addressing other aspects of community assembly and function. For
example, we expect the package will prove useful for i) investigating
how morphological and physiological traits influence growth rate and
shade tolerance, ii) studying emergent phenomena such as multi-species self-thinning, successional transitions and age-related changes in
vegetation productivity, and iii) bridging between empirical data,
large-scale simulators of global biogeochemical cycles and simple
abstract theoretical models.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/patch.pdf}
\caption{Population dynamics for two species competing within a patch.
 Red and blue lines refer to the low- and high-\textsc{lma} species from
 Fig. \ref{fig:plant}. (a) Growth trajectories
 for individuals germinating across a range of patch ages for each
 species, using a schedule of introduction times generated by
 {\plant}'s adaptive algorithm. The initial growth rate advantage of
 the low-\textsc{lma} species (Figure \ref{fig:plant}c) means it
 quickly overtops the high-\textsc{lma} species, suppressing its
 growth. Self-thinning then creates space for the high-\textsc{lma}
 species to establish below the canopy of the faster-growing species.
 (b) Same trajectories as in (a), but shaded
 according to the density $N$ of individuals at given size and patch
 age (light = low density, dark = high density). (c) Leaf area
 index at ground level for each species. The dotted line is the
 total leaf area index (summed across both species), while the lines on
 the right-hand side are averages integrated over the whole
 metapopulation. See Appendix \ref{sec:code} for code
 reproducing this figure.}
\label{fig:patch}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/fitness.pdf}
\caption{Example of a fitness landscape for the \textsc{lma} (leaf mass per
 unit leaf area) trait, showing potential for the stable coexistence of
 multiple plant types \citep[adapted from][]{Falster-2015}. Invasion fitness is the
 logarithm of the basic reproduction ratio --
 at zero fitness, species are at equilibrium, at positive fitness, they increase in
 density. (a) Fitness landscape generated by a single species (indicated by
 a circle). For an \textsc{lma} of 0.2 (dashed line, open circle), there is
 directional selection for smaller \textsc{lma}. For an \textsc{lma} of
  0.0825 (solid line, filled circle), there is a branching point (see also inset). (b) Fitness landscape generated by two species, holding
 the first at an \textsc{lma} of 0.0825. Introducing a species at the
 maximum of the solid line in (a) reduces the region of positive
 fitness considerably (dashed line, open circle). Dotted lines show
 subsequent invasions and replacements of the second species. The
 solid line shows the fitness landscape of the resultant evolutionarily stable two-species community, with both species coexisting
 on separate peaks. At this point, no further
 invasion is possible. See Appendix \ref{sec:code} for code
 reproducing this figure.}
\label{fig:fitness}
\end{figure}
